---
title: "javascript"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{javascript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tabulatoR)
```

## Overview: How tabulatoR Wraps Tabulator.js

tabulatoR provides a lightweight R interface to the powerful [Tabulator](http://tabulator.info/) JavaScript library. Unlike traditional htmlwidgets-based packages, tabulatoR uses a custom Shiny output binding with direct JSON serialization via `jsonlite`, giving you clean, predictable output and full access to Tabulator's extensive API.

### Architecture

The integration consists of three main components:

1. **R functions** ([render.R](../R/render.R)) - `renderTabulatoR()` and `renderSpreadsheet()` serialize your data and configuration to JSON
2. **Custom Shiny binding** ([tabulatoR.js](../inst/tabulatoR.js)) - Handles table creation, updates, and event routing
3. **Proxy functions** ([proxy.R](../R/proxy.R)) - Allow R to invoke methods on existing tables

**Data flow:**

```
R (renderTabulatoR)
  → JSON payload
  → Shiny output binding
  → Creates Tabulator instance
  → User interaction triggers events
  → Events send data back to R via Shiny.setInputValue()
  → R observes input$tableId and reacts
```

### Creating a Table

The main function is `renderTabulatoR()`:

```{r eval=FALSE}
library(shiny)
library(tabulatoR)

ui <- fluidPage(
  tab_source(),  # Load Tabulator CSS/JS from CDN
  tabulatorOutput("myTable")
)

server <- function(input, output, session) {
  output$myTable <- renderTabulatoR(
    mtcars,
    columns = c(
      Column("Model", "model"),
      Column("MPG", "mpg", .editor = "number"),
      Column("Cylinders", "cyl", .editor = "number")
    ),
    autoColumns = FALSE,
    layout = "fitColumns"
  )
}

shinyApp(ui, server)
```

**What happens behind the scenes:**

1. Your data.frame is converted to a list of named lists (object format for JavaScript)
2. Column definitions and options are serialized to JSON
3. The custom output binding receives this payload
4. A new Tabulator instance is created: `new Tabulator(el, options)`
5. The table reference is stored globally: `window[el.id] = table`

### Event Listeners

Event listeners connect JavaScript interactions back to R. tabulatoR provides default handlers for common events, and you can override or add custom ones.

#### Default Event Handlers

These events are automatically registered with sensible defaults:

- **`cellClick`** - Captures: field, value, row data, row index
- **`cellEdited`** - Captures: field, value, old_value, row data, row index
- **`validationFailed`** - Captures: field, value, validation failure info
- **`rowAdded`** - Captures: new row data and index
- **`rowDeleted`** - Captures: deleted row data and index

**Using default events in R:**

```{r eval=FALSE}
server <- function(input, output, session) {
  output$myTable <- renderTabulatoR(mtcars)

  # Listen to default cellEdited event
  observeEvent(input$myTable, {
    event_data <- input$myTable

    cat("Cell edited:\n")
    cat("  Field:", event_data$field, "\n")
    cat("  New value:", event_data$value, "\n")
    cat("  Old value:", event_data$old_value, "\n")
    cat("  Row index:", event_data$index, "\n")
  })
}
```

#### Custom Event Handlers

Override defaults or add new events using the `events` parameter with the `js()` wrapper:

```{r eval=FALSE}
output$myTable <- renderTabulatoR(
  mtcars,
  events = list(
    # Custom cellEdited handler
    cellEdited = js("
      function(cell) {
        return {
          field: cell.getField(),
          newValue: cell.getValue(),
          oldValue: cell.getOldValue(),
          rowIndex: cell.getRow().getPosition(),
          customInfo: 'Modified by user'
        };
      }
    "),

    # Row selection tracking
    rowSelectionChanged = js("
      function(data, rows) {
        return {
          selectedCount: rows.length,
          selectedIds: rows.map(r => r.getData().id)
        };
      }
    "),

    # Cell context menu
    cellContext = js("
      function(e, cell) {
        e.preventDefault();
        return {
          field: cell.getField(),
          value: cell.getValue(),
          position: { x: e.clientX, y: e.clientY }
        };
      }
    ")
  )
)
```

**How it works:**

1. The `js()` function wraps your JavaScript in `<js>...</js>` tags
2. During JSON serialization, this becomes a string
3. The JavaScript binding detects these tags via `parseJSValue()`
4. The string is evaluated back to an executable function: `eval(code)`
5. When the event fires, your function processes the data
6. The returned object is sent to R via `Shiny.setInputValue()`

#### Available Tabulator Events

You can register handlers for any [Tabulator event](http://tabulator.info/docs/5.5/events). Common ones include:

- **Data events:** `dataLoaded`, `dataChanged`, `dataFiltered`, `dataSorted`
- **Cell events:** `cellClick`, `cellDblClick`, `cellEdited`, `cellMouseEnter`
- **Row events:** `rowClick`, `rowAdded`, `rowDeleted`, `rowMoved`, `rowSelected`
- **Column events:** `columnMoved`, `columnResized`, `columnVisibilityChanged`
- **Validation:** `validationFailed`
- **Editing:** `dataEdited`, `editableCheck`

### Invoking Methods from R (Proxy Functions)

Proxy functions let R code modify an existing table without re-rendering. They use `session$sendCustomMessage()` to communicate with JavaScript.

#### Adding Rows

```{r eval=FALSE}
server <- function(input, output, session) {
  output$myTable <- renderTabulatoR(mtcars)

  observeEvent(input$addRowBtn, {
    new_row <- data.frame(
      model = "New Car",
      mpg = 25,
      cyl = 4,
      stringsAsFactors = FALSE
    )

    # Add to bottom (default)
    tabulatorAddRow("myTable", new_row)

    # Or add to top
    tabulatorAddRow("myTable", new_row, add_to = "top")
  })
}
```

**Behind the scenes:**
```
R: tabulatorAddRow("myTable", data, "bottom")
  → session$sendCustomMessage("tabulator-add-data", list(id = "myTable", data = data, add_to = "bottom"))
  → JS: Shiny.addCustomMessageHandler receives message
  → Finds table: document.getElementById("myTable")
  → Calls: table.addData(message.data, message.add_to)
```

#### Removing Rows

```{r eval=FALSE}
# Remove row at index 5
tabulatorRemoveRow("myTable", index = 5)
```

#### Replacing All Data

```{r eval=FALSE}
# Replace entire table with new data
tabulatorReplaceData("myTable", new_data_frame)
```

#### Reverting Edits

```{r eval=FALSE}
observeEvent(input$myTable, {
  edited <- input$myTable

  # Validation check
  if (edited$value < 0) {
    showNotification("Negative values not allowed!", type = "error")

    # Revert the cell to its previous value
    tabulatorRevertField(
      "myTable",
      index = edited$index,
      field = edited$field
    )
  }
})
```

#### Spreadsheet-Specific Proxies

For tables created with `renderSpreadsheet()`:

```{r eval=FALSE}
# Replace spreadsheet data
spreadsheetSetData("mySheet", new_data)

# Clear all cells
spreadsheetClearSheet("mySheet")

# Get current spreadsheet data (triggers input$mySheet_data)
spreadsheetGetData("mySheet")

observeEvent(input$mySheet_data, {
  current_data <- input$mySheet_data
  # Process or save the data
})
```

### Direct Tabulator API Access

For advanced use cases, you can access the Tabulator instance directly from JavaScript:

```{r eval=FALSE}
output$myTable <- renderTabulatoR(
  mtcars,
  .opts = list(
    # Any Tabulator option works here
    pagination = TRUE,
    paginationSize = 10,
    groupBy = "cyl",
    movableRows = TRUE,
    selectable = "highlight"
  )
)
```

Or use browser console JavaScript:

```javascript
// Access the table (if table id is "myTable")
const table = window.myTable;

// Call any Tabulator method
table.deselectRow();
table.setFilter("mpg", ">", 20);
table.download("csv", "data.csv");
```

### Complete Example: Editable Table with Validation

```{r eval=FALSE}
ui <- fluidPage(
  tab_source(theme = "tabulator_bootstrap5"),
  actionButton("addRow", "Add Row"),
  actionButton("getData", "Get Current Data"),
  tabulatorOutput("editableTable"),
  verbatimTextOutput("debugInfo")
)

server <- function(input, output, session) {
  # Reactive to store current valid data
  table_data <- reactiveVal(mtcars[1:5, ])

  output$editableTable <- renderTabulatoR(
    table_data(),
    columns = c(
      Column("Model", "model"),
      Column("MPG", "mpg", .editor = "number", .validator = "min:0"),
      Column("Cylinders", "cyl", .editor = "list",
             .editorParams = list(values = c(4, 6, 8)))
    ),
    events = list(
      cellEdited = js("
        function(cell) {
          return {
            field: cell.getField(),
            value: cell.getValue(),
            old_value: cell.getOldValue(),
            index: cell.getRow().getPosition()
          };
        }
      "),
      validationFailed = js("
        function(cell, value, validators) {
          return {
            field: cell.getField(),
            invalidValue: value,
            validators: validators
          };
        }
      ")
    ),
    layout = "fitColumns"
  )

  # Handle valid edits
  observeEvent(input$editableTable, {
    edit <- input$editableTable

    # Update our reactive data store
    current <- table_data()
    current[edit$index, edit$field] <- edit$value
    table_data(current)

    output$debugInfo <- renderPrint({
      cat("Valid edit:\n")
      cat("  Row:", edit$index, "\n")
      cat("  Field:", edit$field, "\n")
      cat("  Old:", edit$old_value, "\n")
      cat("  New:", edit$value, "\n")
    })
  })

  # Handle validation failures
  observeEvent(input$editableTable_validationFailed, {
    fail <- input$editableTable_validationFailed

    showNotification(
      paste0("Invalid value for ", fail$field, ": ", fail$invalidValue),
      type = "error"
    )

    # Revert to old value
    tabulatorRevertField(
      "editableTable",
      index = fail$index,
      field = fail$field
    )
  })

  # Add new row
  observeEvent(input$addRow, {
    new_row <- data.frame(
      model = "New Car",
      mpg = 20,
      cyl = 4,
      stringsAsFactors = FALSE
    )
    tabulatorAddRow("editableTable", new_row, add_to = "bottom")
  })
}

shinyApp(ui, server)
```

This example demonstrates the full lifecycle: initial render, event handling, validation, and proxy-based table manipulation.
